The \Rpackage{eisa} and \Rpackage{biclust} packages

### 2009-12-08 eisa/inst/doc/EISA_biclust.Rnw
eisa vignettes updated.

The Iterative Signature Algorithm (ISA)~\cite{isa} is a biclustering
method, that can efficiently find potentially overlapping biclusters
(modules, according to the ISA terminology) in a matrix. The ISA is
implemented in the \Rpackage{eisa} package~\cite{eisapackage} it uses
standard BioConductor classes and includes a number of visualization
tools as well.

@@

Next, we select a subset of the genes in the data set. We do this to speed 
up the computation for our simple examples. We select the genes that are 
annotated to involved in immune system processes, according to the Gene
Ontology database.

@@

Now we will convert the \Rclass{Biclust} object to an \Rclass{ISAModules} 
object, that is used in the \Rpackage{eisa} package. To help
some \Rpackage{eisa} functions, we add the name of the annotation package to 
the parameters stored in the \Rclass{Biclust} object, this is always adwised.
The procedure makes use of the probe and sample names that are kept and stored 
in the \Rclass{Biclust} object, this information will be used later, e.g. for 
the enrichment analysis. The conversion itself can be performed with
the usual \Rfunction{as} function. 
<<convert bc to isa>>=
Bc@Parameters$annotation <- annotation(ALL.filtered)
modules <- as(Bc, "ISAModules")
modules
@

@@

Doing some enrichment analysis is easy:
<<biclust enrichment>>=
library(KEGG.db)
KEGG <- ISAKEGG(modules)
sigCategories(KEGG)[[2]]
unlist(mget(sigCategories(KEGG)[[2]], KEGGPATHID2NAME))
@

@@

\begin{figure}
\centering
<<heatmap-real,fig=TRUE,echo=FALSE>>=
<<heatmap>>
@
\caption{Heatmap of the second module, found with the Plaid Model
  biclustering algorithm. The black squares denote the T-cell samples;
  all samples in the module belong to T-cell samples.}
\label{fig:heatmap}
\end{figure}

@@

@@

@@

@@

@@

The conversion from \Rclass{ISAModules} to \Rclass{Biclust} can be done
the usual way:
<<isa to biclust>>=
BcMods <- as(ALLModules, "Biclust")
BcMods
@

@@

Let's create a pairs-plot to visualize the relationship of these measures
for our data set, the result is in Fig.~\ref{fig:coherence}.
<<pairs,eval=FALSE>>=
pairs( cbind(cV, aV, mV, sV, rob) )
@
\begin{figure}
\centering
<<pairs-real,fig=TRUE,echo=FALSE>>=
<<pairs>>
@ 
\caption{Relationship of the various bicluster coherence measueres and
  the ISA robustness measure. They show high correlation.}
\label{fig:coherence}
\end{figure}

>>>>

The Iterative Signature Algorithm (ISA)~\cite{sa,isa,isamod} is a
biclustering method, that can efficiently find potentially overlapping
biclusters (modules, according to the ISA terminology) in a
matrix. The ISA is implemented in the \Rpackage{eisa} package. This
package uses standard BioConductor classes and includes a number of
visualization tools as well.

@@

Next, we select a subset of the genes in the data set. We do this to speed 
up the computation for our simple examples. We select the genes that are 
annotated to be involved in immune system processes, according to the Gene
Ontology database.

@@

Now we will convert the \Rclass{Biclust} object to an \Rclass{ISAModules} 
object, that is used in the \Rpackage{eisa} package. To help
some \Rpackage{eisa} functions, we add the name of the annotation package to 
the parameters stored in the \Rclass{Biclust} object, this is always advised.
The procedure makes use of the probe and sample names that are kept and stored 
in the \Rclass{Biclust} object, this information will be used later, e.g. for 
the enrichment analysis. The conversion itself can be performed with
the usual \Rfunction{as} function. 
<<convert bc to isa>>=
Bc@Parameters$annotation <- annotation(ALL.filtered)
modules <- as(Bc, "ISAModules")
modules
@

\subsection{Enrichment analysis}

@@

Performing enrichment analysis is easy:
<<biclust enrichment>>=
library(KEGG.db)
KEGG <- ISAKEGG(modules)
sigCategories(KEGG)[[2]]
unlist(mget(sigCategories(KEGG)[[2]], KEGGPATHID2NAME))
@

\subsection{Heatmaps}

@@

\begin{figure}
\centering
<<heatmap-real,fig=TRUE,echo=FALSE>>=
<<heatmap>>
@
\caption{Heatmap of the second module, found with the Plaid Model
  biclustering algorithm. The black squares denote the T-cell samples;
  all samples in the module are from T-cell leukemia patients.}
\label{fig:heatmap}
\end{figure}

@@

\subsection{Profile plots}

@@

\subsection{Gene Ontology tree plots}

@@

\subsection{HTML summary of the biclusters}

@@

\subsection{Group-mean plots}

@@

The conversion from \Rclass{ISAModules} to \Rclass{Biclust} can be done
the usual way, using the \Rfunction{as} function:
<<isa to biclust>>=
BcMods <- as(ALLModules, "Biclust")
BcMods
@

\subsection{Coherence of biclusters}

@@

Let's create a pairs-plot to visualize the relationship of these measures
for our data set, the result is in Fig.~\ref{fig:coherence}.
<<pairs,eval=FALSE>>=
panel.low <- function(x, y) {
  usr <- par("usr")
  m <- c((usr[2]+usr[1])/2, (usr[4]+usr[3])/2)
  text(m[1], m[2], adj=c(1/2,1/2), cex=1.5,
       paste(sep="\n", "Correlation:", round(cor(x,y),2)))
}
pairs( cbind(cV, aV, mV, sV, rob), lower.panel=panel.low )
@
\begin{figure}
\centering
<<pairs-real,fig=TRUE,echo=FALSE>>=
<<pairs>>
@ 
\caption{Relationship of the various bicluster coherence measueres and
  the ISA robustness measure. They show high correlation.}
\label{fig:coherence}
\end{figure}
